---
title: "Milestone Report - Coursera Data Science Swiftkey Capstone Project"
author: "Ricky Purnomo"
date: "Sunday, November 16, 2014"
output: html_document
---

```{r setoptions, echo=FALSE}
opts_chunk$set(echo = FALSE, results = "hide")
```

This report aims to illustrate the foundation in preparation to create a word prediction algorithm as part of the Coursera Data Science capstone project. 

The motivation for this project is to: 

1. Demonstrate that you've downloaded the data and have successfully loaded it in.
2. Create a basic report of summary statistics about the data sets.
3. Report any interesting findings that you amassed so far.
4. Get feedback on your plans for creating a prediction algorithm and Shiny app. 

## 

```{r}
summary(cars)
```

The number of words available from the collection for prediction is huge and may be too much for quick processing. Fortunately [ bla bla ] As shown below, a small number of unique terms are needed to represent majority of the words in a whole set.

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
